---
title: "Intermediate Elective 2"
author: "Rebecca Martinez"
date: "15 May 2026"
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

# 1. Inspiration

## Visualization inspiration

I chose Steven Ponce’s “Coal falls. Renewables rise.” as my inspiration. I liked how the visualization uses two directly labeled time series lines, a dark background, and contrasting colors to show two categories changing through time. I also liked that the color contrast helps separate a less desirable trend from a more desirable one, which fits my goal of comparing non-native and native vegetation cover in vernal pools.

## Why it makes sense for my dataset

This inspiration makes sense for my dataset because the vernal pool vegetation data include percent cover values across multiple years. Since I am interested in comparing native and non-native vegetation cover through time, a two-line time series is a good way to show whether these groups follow similar or contrasting patterns.

## Variables and visual mapping

The x-axis will show year, and the y-axis will show mean percent cover across vernal pools. The `cover_group` variable will be mapped to color so native and non-native vegetation are shown as separate lines. Points will show yearly values, lines will show the trend through time, and direct labels/annotations will help emphasize the contrast between native and non-native cover.

# 2. Planned figure

The sketch below shows the planned layout for my figure. It includes two time series lines, direct labels for native and non-native cover, and a dark background inspired by the original visualization.

[figure placeholder]


# 3. Code the figure

## Setup

### Packages

```{r packages}

# general use
library(tidyverse) # data cleaning, wrangling, and plotting
library(janitor) # clean column names
library(glue) # combine text and values for labels
library(scales) # format axis labels

# file organization
library(here) # create reproducible file paths

# data visualization
library(naniar) # visualize missing data
library(patchwork) # arrange plots
library(viridis) # color palette
library(ggtext) # styled text in ggplot titles/subtitles


```

### Data

```{r data}

# read in vegetation data
veg <- read_csv(here("data", "veg.csv"))

# read in vernal pool vegetation metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))


```

## Cleaning

### Create explicit 0s

```{r explicit 0s}

# create explicit-zero data frame by species
pools_explicit0 <- veg |> 
  
  # keep only vernal pool transects at NCOS
  filter(str_detect(transect_name, "VP") &
           site == "ncos") |> 
  
  # add missing combinations and fill missing percent cover with 0
  complete(year,
           transect_name,
           psoc,
           transect_distance,
           fill = list(percent_cover = 0)) |> 
  
  # group by year, transect, and species
  group_by(year, transect_name, psoc) |> 
  
  # calculate total percent cover
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE),
            .groups = "drop") |> 
  
  # create a shared column for joining with metadata
  # keep year and transect_name because we need them later
  unite("year_pool",
        year,
        transect_name,
        remove = FALSE) |> 
  
  # join metadata to add number of quadrats
  left_join(metadata |>
              select(year_pool, num_quad),
            by = "year_pool") |> 
  
  # calculate mean percent cover using number of quadrats
  mutate(mean_pc = sum_pc / num_quad)

# check cleaned data
pools_explicit0 |>
  slice_sample(n = 10)

# check column names
names(pools_explicit0)

# check missingness after cleaning
gg_miss_var(pools_explicit0)


```

### Create native and non-native vectors

```{r native and non-native vectors}


# create a vector of native species names
native_spp <- veg |>
  
  # keep only native cover rows
  filter(cover_category == "NATIVE COVER") |>
  
  # pull species column as a vector
  pull(psoc) |>
  
  # keep only unique species names
  unique()


# create a vector of non-native species names
nonnative_spp <- veg |>
  
  # keep only non-native cover rows
  filter(cover_category == "NON-NATIVE COVER") |>
  
  # pull species column as a vector
  pull(psoc) |>
  
  # keep only unique species names
  unique()

# check number of species in each group
length(native_spp)
length(nonnative_spp)

```

### Add cover vectors to explicit data

```{r add cover vectors to explicit data}

# add native/non-native grouping to explicit-zero data
cover_explicit0 <- pools_explicit0 |>
  
  # create cover group based on species vectors
  mutate(cover_group = case_when(
    psoc %in% native_spp ~ "Native",
    psoc %in% nonnative_spp ~ "Non-native",
    TRUE ~ NA_character_
  )) |>
  
  # remove species that were not classified as native or non-native
  filter(!is.na(cover_group)) |>
  
  # keep columns needed for summarizing and plotting
  select(year,
         transect_name,
         psoc,
         cover_group,
         sum_pc,
         num_quad,
         mean_pc)

# check grouped data
cover_explicit0 |>
  slice_sample(n = 10)

```


### Summarize cover through time

```{r summarize data}

# summarize native and non-native cover by year, pool, and cover group
cover_pool_year <- cover_explicit0 |>
  
  # group by year, pool, and cover group
  group_by(year, transect_name, cover_group) |>
  
  # calculate total mean percent cover across species
  summarize(total_mean_pc = sum(mean_pc, na.rm = TRUE),
            .groups = "drop")


# summarize average native and non-native cover across pools by year
cover_year_summary <- cover_pool_year |>
  
  # group by year and cover group
  group_by(year, cover_group) |>
  
  # calculate average cover across pools
  summarize(mean_cover = mean(total_mean_pc, na.rm = TRUE),
            sd_cover = sd(total_mean_pc, na.rm = TRUE),
            n_pools = n_distinct(transect_name),
            .groups = "drop")

# display summarized data
cover_year_summary


```

## Final figure

```{r final figure}

# create endpoint labels for each line
end_labels <- cover_year_summary |>
  
  # keep the last year for each cover group
  group_by(cover_group) |>
  
  # keep the row with the largest year
  filter(year == max(year)) |>
  
  # ungroup data frame
  ungroup() |>
  
  # create label with cover group and rounded percent cover
  mutate(label = glue("{cover_group}\n{round(mean_cover, 1)}%"))


# set plot colors inspired by Steven Ponce's visualization
bg_color <- "#0D1B2A"
native_gold <- "#F4A932"
nonnative_blue <- "#6A80A0"
text_white <- "#FFFFFF"
text_muted <- "#8A98A8"
text_sub <- "#B0B8C4"
grid_color <- "#1A2E42"


# create styled title with colored text
title_text <- glue(
  "<span style='font-weight:900; color:{native_gold};'>Native cover remains dominant.&nbsp;</span>",
  "<span style='font-weight:900; color:{nonnative_blue};'> Non-native cover persists.</span>"
)


# create styled subtitle
subtitle_text <- glue(
  "<span style='font-size:12pt; color:{text_sub};'>",
  "Native and non-native vegetation cover across NCOS vernal pools<br>",
  "Mean percent cover · 2021–2025",
  "</span>"
)


# create final plot inspired by Steven Ponce's visualization
final_cover_plot <- ggplot(data = cover_year_summary,
                           mapping = aes(x = year,
                                         y = mean_cover,
                                         color = cover_group,
                                         group = cover_group)) +
  
  # add horizontal grid lines
  geom_hline(yintercept = seq(0, 70, by = 10),
             color = grid_color,
             linewidth = 0.25) +
  
  # add native line with stronger emphasis
  geom_line(data = cover_year_summary |> 
              filter(cover_group == "Native"),
            linewidth = 1.9,
            alpha = 1) +
  
  # add non-native line with softer emphasis
  geom_line(data = cover_year_summary |> 
              filter(cover_group == "Non-native"),
            linewidth = 1.3,
            alpha = 0.70) +
  
  # add points for the first and last years
  geom_point(data = cover_year_summary |>
               group_by(cover_group) |>
               filter(year == min(year) | year == max(year)) |>
               ungroup(),
             size = 3.5,
             shape = 21,
             fill = bg_color,
             stroke = 1.5) +
  
  # add direct labels at the end of each line
  geom_text(data = end_labels,
            mapping = aes(label = label),
            x = max(cover_year_summary$year) + 0.25,
            hjust = 0,
            vjust = 0.5,
            size = 3.2,
            lineheight = 1.1,
            show.legend = FALSE) +
  
  # set native and non-native colors
  scale_color_manual(values = c("Native" = native_gold,
                                "Non-native" = nonnative_blue)) +
  
  # set x-axis breaks and add space for direct labels
  scale_x_continuous(breaks = sort(unique(cover_year_summary$year)),
                     expand = expansion(mult = c(0.03, 0.24))) +
  
  # format y-axis as percent cover
  scale_y_continuous(limits = c(0, 70),
                     breaks = seq(0, 70, by = 10),
                     labels = \(x) glue("{x}%"),
                     expand = expansion(mult = c(0.01, 0.04))) +
  
  # add title, subtitle, and caption
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "Data: Vernal pool vegetation monitoring data",
       x = NULL,
       y = NULL) +
  
  # use a dark theme similar to the inspiration figure
  theme_void() +
  
  # customize theme elements
  theme(
    plot.background = element_rect(fill = bg_color, color = NA),
    panel.background = element_rect(fill = bg_color, color = NA),
    
    plot.title = element_markdown(color = text_white,
                                  size = 24,
                                  hjust = 0,
                                  lineheight = 1.1,
                                  margin = margin(t = 10, b = 6)),
    
    plot.subtitle = element_markdown(color = text_sub,
                                     size = 10,
                                     hjust = 0,
                                     lineheight = 1.4,
                                     margin = margin(b = 22)),
    
    plot.caption = element_text(color = text_muted,
                                size = 7,
                                hjust = 1,
                                margin = margin(t = 14)),
    
    axis.text.x = element_text(size = 9,
                               color = text_muted,
                               margin = margin(t = 6)),
    
    axis.text.y = element_text(size = 8,
                               color = text_muted,
                               hjust = 1),
    
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 20, r = 55, b = 16, l = 20)
  ) +
  
  # allow labels to extend beyond plot area
  coord_cartesian(clip = "off")

# display final plot
final_cover_plot




```


