---
title: "Intermediate Elective 2"
author: "Rebecca Martinez"
date: "15 May 2026"
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

# 1. Inspiration

## Visualization inspiration

I chose Steven Ponce’s “Coal falls. Renewables rise.” as my inspiration. I liked how the visualization uses two directly labeled time series lines, a dark background, and contrasting colors to show two categories changing through time. I also liked that the color contrast helps separate a less desirable trend from a more desirable one, which fits my goal of comparing non-native and native vegetation cover in vernal pools.

## Why it makes sense for my dataset

This inspiration makes sense for my dataset because the vernal pool vegetation data include percent cover values across multiple years. Since I am interested in comparing native and non-native vegetation cover through time, a two-line time series is a good way to show whether these groups follow similar or contrasting patterns.

## Variables and visual mapping

The x-axis will show year, and the y-axis will show mean percent cover across vernal pools. The `cover_group` variable will be mapped to color so native and non-native vegetation are shown as separate lines. Points will show yearly values, lines will show the trend through time, and direct labels/annotations will help emphasize the contrast between native and non-native cover.

# 2. Planned figure

The sketch below shows the planned layout for my figure. It includes two time series lines, direct labels for native and non-native cover, and a dark background inspired by the original visualization.

[figure placeholder]


# 3. Code the figure

## Setup

### Packages

```{r packages}

# general use
library(tidyverse) # data cleaning, wrangling, and plotting
library(janitor) # clean column names
library(glue) # combine text and values for labels
library(scales) # format axis labels
library(lubridate) # clean dates

# file organization
library(here) # create reproducible file paths

# data visualization
library(naniar) # visualize missing data
library(patchwork) # arrange plots
library(viridis) # color palette
library(ggtext) # styled text in ggplot titles/subtitles


```

### Data

```{r data}

# read in vegetation data
veg <- read_csv(here("data", "veg.csv"))

# read in vernal pool vegetation metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# read in NOAA weather data for possible precipitation annotations
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

```

## Cleaning

### Clean vegetation data

```{r clean veg data}

# create observed percent cover data frame by species and sampling month
pools_observed <- veg |> 
  
  # keep only vernal pool transects at NCOS
  filter(str_detect(transect_name, "VP") &
           site == "ncos") |>
  
  # keep only columns needed for this analysis
  select(date,
         year,
         transect_name,
         psoc,
         cover_category,
         percent_cover) |>
  
  # remove rows with missing percent cover
  filter(!is.na(percent_cover)) |>
  
  # create month column from sampling date
  mutate(month = month(date)) |>
  
  # group by year, month, transect, and species
  group_by(year, month, transect_name, psoc) |>
  
  # calculate mean percent cover from observed records
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE),
            .groups = "drop")

# check cleaned data
pools_observed |>
  slice_sample(n = 10)

```

### Create native and non-native vectors

```{r native and non-native vectors}

# create a vector of native species names
native_spp <- veg |>
  
  # keep only native cover rows
  filter(cover_category == "NATIVE COVER") |>
  
  # pull species column as a vector
  pull(psoc) |>
  
  # keep only unique species names
  unique()


# create a vector of non-native species names
nonnative_spp <- veg |>
  
  # keep only non-native cover rows
  filter(cover_category == "NON-NATIVE COVER") |>
  
  # pull species column as a vector
  pull(psoc) |>
  
  # keep only unique species names
  unique()

# check number of species in each group
length(native_spp)
length(nonnative_spp)

```

### Add cover vectors to explicit data

```{r add cover vectors to data}

# add native/non-native grouping to observed data
cover_observed <- pools_observed |>
  
  # create cover group based on species vectors
  mutate(cover_group = case_when(
    psoc %in% native_spp ~ "Native",
    psoc %in% nonnative_spp ~ "Non-native",
    TRUE ~ NA_character_
  )) |>
  
  # remove species that were not classified as native or non-native
  filter(!is.na(cover_group))

# check grouped data
cover_observed |>
  slice_sample(n = 10)

```


### Summarize cover by month and year

```{r summarize by month  and year}

### Summarize cover by month

# summarize native and non-native cover by month, pool, and cover group
cover_pool_month <- cover_observed |>
  
  # group by year, month, pool, and cover group
  group_by(year, month, transect_name, cover_group) |>
  
  # calculate total mean percent cover across observed species
  summarize(total_mean_pc = sum(mean_pc, na.rm = TRUE),
            .groups = "drop")


# create month/year order for plotting
month_order <- cover_pool_month |>
  
  # keep unique year/month combinations
  distinct(year, month) |>
  
  # arrange in time order
  arrange(year, month) |>
  
  # create numeric time index and readable month/year label
  mutate(time_index = row_number(),
         month_year = glue("{month.abb[month]} {year}"))


# summarize average native and non-native cover across pools by month
cover_month_summary <- cover_pool_month |>
  
  # group by year, month, and cover group
  group_by(year, month, cover_group) |>
  
  # calculate average cover across vernal pool transects
  summarize(mean_cover = mean(total_mean_pc, na.rm = TRUE),
            sd_cover = sd(total_mean_pc, na.rm = TRUE),
            n_pools = n_distinct(transect_name),
            .groups = "drop") |>
  
  # join month/year order for plotting
  left_join(month_order, by = c("year", "month"))

# display summarized data
cover_month_summary

```

### Clean weather data

```{r}

# clean NOAA precipitation data
weather_clean <- weather |>
  
  # clean column names
  clean_names() |>
  
  # change date column to standard date format
  mutate(date = mdy(date)) |>
  
  # keep precipitation column and create year/month columns
  select(date, prcp) |>
  mutate(year = year(date),
         month = month(date),
         plot_date = make_date(year = year,
                               month = month,
                               day = 1))

```

### Summarize precipitation by month

```{r}

# summarize total precipitation by year and month
monthly_precip <- weather_clean |>
  
  # keep years that overlap with vegetation data
  filter(year %in% unique(cover_month_summary$year)) |>
  
  # group by year, month, and plot date
  group_by(year, month, plot_date) |>
  
  # calculate total monthly precipitation
  summarize(total_precip = sum(prcp, na.rm = TRUE),
            .groups = "drop")

# identify the wettest month in each year
wettest_months <- monthly_precip |>
  
  # group by year
  group_by(year) |>
  
  # keep the month with the highest precipitation in each year
  slice_max(total_precip, n = 1, with_ties = FALSE) |>
  
  # ungroup data frame
  ungroup()

# display wettest months
wettest_months

```


## Final figure
